Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}

### 2020-08-06 penguins.Rmd
add bookdown to enable cross references in text

---
title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's *Irises*
author:
  - name: Allison M. Horst
    affiliation: Bren School of Environmental Science and Management
    address:
    - University of California, Santa Barbara
    - Santa Barbara, CA 93106-5131
    email:  ahorst@ucsb.edu
  - name: Alison P. Hill
    affiliation: RStudio, PBC
    address:
    - 250 Northern Ave
    - Boston, MA 02210
    email:  alison@rstudio.com
  - name: Kristen B. Gorman
    affiliation: University of Alaska Fairbanks College of Fisheries and Ocean Sciences
    address:
    - 2150 Koyukuk Drive
    - 245 O’Neill Building
    - Fairbanks, AK 99775-7220
    email:  kbgorman@alaska.edu
abstract: >

 The famous *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education. While the **iris** measurements were originally collected by botanist Edgar Anderson, their use in Ronald A. Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace **iris** with alternate data in teaching materials. Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data. Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data. We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g. PCA), cluster analysis and classification (e.g. by k-means).

preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
---

>>>>

---
title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative
  to Anderson's *Irises*
author:
- name: Allison M. Horst
  affiliation: Bren School of Environmental Science and Management
  address:
  - University of California, Santa Barbara
  - Santa Barbara, CA 93106-5131
  email: ahorst@ucsb.edu
- name: Alison P. Hill
  affiliation: RStudio, PBC
  address:
  - 250 Northern Ave
  - Boston, MA 02210
  email: alison@rstudio.com
- name: Kristen B. Gorman
  affiliation: University of Alaska Fairbanks College of Fisheries and Ocean Sciences
  address:
  - 2150 Koyukuk Drive
  - 245 O’Neill Building
  - Fairbanks, AK 99775-7220
  email: kbgorman@alaska.edu
abstract: "\nThe famous *Iris* dataset, which exists in the base R **datasets** package
  as **iris**, is ubiquitous in statistics and data science education. While the **iris**
  measurements were originally collected by botanist Edgar Anderson, their use in
  Ronald A. Fisher's eugenics research has created momentum to address the dataset's
  racist use and/or replace **iris** with alternate data in teaching materials. Penguin
  morphological measurements for three species of *Pygoscelis* penguins that breed
  on islands throughout the Palmer Archipelago, Antarctica, provide an approachable,
  charismatic and near drop-in replacement for the **iris** data. Here, we introduce
  the **palmerpenguins** R package and describe the included **penguins** data. We
  directly compare the two datasets for selected analyses to demonstrate that R users,
  in particular teachers and learners currently using the **iris** data, can switch
  to the Palmer Archipelago penguins for many use cases including data wrangling and
  visualization, correlation, regression, hypothesis testing, multivariate analysis
  (e.g. PCA), cluster analysis and classification (e.g. by k-means). \n"
output:
  bookdown::pdf_book:
    base_format: rticles::rjournal_article
preamble: |
  % Any extra LaTeX you need in the preamble
---

### 2020-08-06 penguins.tex
add bookdown to enable cross references in text

\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

@@

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
\end{figure}
\end{Schunk}

@@

\caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B). Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
\end{figure}
\end{Schunk}

@@

\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
\end{figure}
\end{Schunk}

@@

\caption[Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component]{Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component. For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares). For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares). Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}\label{fig:pca}
\end{figure}
\end{Schunk}

@@

\caption[K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B)]{K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B). Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.}\label{fig:kmeans}
\end{figure}
\end{Schunk}

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
\hline
2 & 4 & 54 & 6 & 2 & NA & 2 & 46\\
\hline
3 & 147 & 5 & 1 & 3 & 50 & NA & NA\\
\hline
\end{tabular}
\end{table}

>>>>

\caption{(\#tab:unnamed-chunk-1)Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

@@

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}(\#fig:penguins-pairs)
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}(\#fig:iris-pairs)
\end{figure}
\end{Schunk}

@@

\caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B). Within-species linear model is visualized for each penguin or iris species.}(\#fig:linear-example)
\end{figure}
\end{Schunk}

@@

\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}(\#fig:simpsons)
\end{figure}
\end{Schunk}

@@

\caption[Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component]{Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component. For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares). For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares). Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}(\#fig:pca)
\end{figure}
\end{Schunk}

@@

\caption[K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B)]{K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B). Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.}(\#fig:kmeans)
\end{figure}
\end{Schunk}

@@

\caption{(\#tab:unnamed-chunk-3)K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
\hline
2 & NA & 9 & 116 & 2 & NA & 2 & 46\\
\hline
3 & 147 & 5 & 1 & 3 & NA & 48 & 4\\
\hline
\end{tabular}
\end{table}
